---
title: "Our question!"
format: html
editor: visual
execute:
  warning: false
---

## Introduction

Here we aim to answer the question: "Does rainfall correlate with change in K & NO3 at Bisley measurement sites?" This report uses three different sites at Bisley: Quebrada Cuenca 1, Quebrada Cuenca 2, and Quebrada Cuenca 3, as well as rainfall data for the region, highlighting the time when Hurricane Hugo hit in 1989.

## Load in the packages

```{r}
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(patchwork)
source(here("R", "convert_monthly_to_bimonthly.R"))
```

## Read in the four files needed to analyze the Bisley sites vs. rainfall

```{r}
Q1 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca1-Bisley.csv")) %>%
  clean_names()

Q2 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca2-Bisley.csv")) %>%
  clean_names()

Q3 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca3-Bisley.csv")) %>%
  clean_names()

rainfall <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "BisleywklyRain-Throughfall1988-2015.csv")) %>%
  clean_names()
```

## Combine chemistry datasets by a full join

```{r}
bisley_sites_joined_df <- Q1 %>%
  full_join(Q2) %>%
  full_join(Q3)
```

## Tidy the data and begin to make it ready for our analysis

```{r}
# Chemistry dataframe
bisley_pivot_df_no3_k <- bisley_sites_joined_df %>%
  select(sample_id, sample_date, no3_n, k) %>% # Select only necessary columns
  filter(lubridate::year(sample_date) %in% 1989:1994) %>% # Only have data from 1989-1994
  mutate(month_date = lubridate::floor_date(sample_date, "month")) %>% # Make data from the day level to the month level
  pivot_longer(cols = c(no3_n, k), names_to = "chemicals", values_to = "amounts") # Make the two chemical names in a column and values in another

# Rainfall dataframe
rainfall_subset <- rainfall %>%
  mutate(date = lubridate::mdy(date)) %>% # Put dates in same format as other dataframe
  select(date, rammday) %>% # Don't need throughfall column
  filter(lubridate::year(date) %in% 1989:1994) %>% # Only have data from 1989-1994
  mutate(month_date = lubridate::floor_date(date, "month"), .after = date) # Make data from the day level to the month level
```

## Add a column with the level of the data as bimonthly (every two months) rather than monthly

```{r}
# For chemistry data
bisley_pivot_df_no3_k <- convert_monthly_to_bimonthly(df = bisley_pivot_df_no3_k, month_column = month_date)

# For rainfall data
rainfall_subset <- convert_monthly_to_bimonthly(df = rainfall_subset, month_column = month_date)
```

## Create a summary table with means by month, chemical, and site

```{r}
# Summarize chemistry df means by month, chemical, and site
chemical_summary_df <- bisley_pivot_df_no3_k %>%
  group_by(sample_id, odd_month_date, chemicals) %>%
  summarize(chemical_means_by_month = mean(amounts, na.rm = TRUE))

# Summarize rainfall df means by month
rainfall_summary_df <- rainfall_subset %>%
  group_by(odd_month_date) %>%
  summarize(rainfall_means_by_month = mean(rammday, na.rm = TRUE))
```

## Create a plot of the two chemicals of interest (K and NO3) with colors by site with another plot containing rainfall for the same time period

```{r}
# Define positions of vline for Hugo
chemical_date_vline <- as.Date(c("1989-09-01"))    
chemical_date_vline_position <- which(chemical_summary_df$odd_month_date %in% chemical_date_vline)

rainfall_date_vline <- as.Date(c("1989-09-01"))    
rainfall_vline_position <- which(rainfall_summary_df$odd_month_date %in% rainfall_date_vline)

# Plot the two chemicals of interest with colors by site
chemical_plot <- ggplot(data = chemical_summary_df, aes(x = odd_month_date, 
                              y = chemical_means_by_month)) +
  geom_line(aes(color = sample_id)) +
  facet_wrap(~chemicals,
             scales = "free_y",
             ncol = 1,
             strip.position = "left", 
             labeller = as_labeller(c(k = "K mg l^-1", 
                                      no3_n = "NO3-N ug l^-1"))) +
  scale_color_manual(values = c("darkslategray3", "darkslategray4", "darkslategray")) +
  labs(x = "", y = "Bimonthly Chemical Means", color = "") +
  theme_light() +
  geom_vline(xintercept = as.numeric(chemical_summary_df$odd_month_date[chemical_date_vline_position]), 
             color = "dark grey",
             linetype = "dashed") + # Add the line
  theme(panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank(), # Get rid of vertical grid lines 
        text=element_text(size=10,  family="Times New Roman")) 

# Plot rainfall
rainfall_plot <- ggplot(data = rainfall_summary_df, aes(x = odd_month_date, 
                              y = rainfall_means_by_month)) +
  geom_line(color = "darkblue") +
  labs(x = "Years", y = "Bimonthly Rainfall Means (mm)") +
  theme_light() + 
  geom_vline(xintercept = as.numeric(chemical_summary_df$odd_month_date[chemical_date_vline_position]), 
             color = "dark grey",
             linetype = "dashed") + # Add the line
  theme(panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank(), # Get rid of vertical grid lines 
        text=element_text(size=10,  family="Times New Roman")) # Change text size and font

# Create combined plot
combined_plot <- chemical_plot / rainfall_plot
combined_plot
```

## Save the plot to the plots folder

```{r}
ggsave(plot = combined_plot,
       here("plots", "rainfall_and_stream_chemistry_plot.png"),
       width = 8,
       height = 8,
       dpi = 200)
```
