---
title: "Our recreated plot!"
format: html
editor: visual
execute:
  warning: false
---

## Introduction

Here we aim to recreate a previously-made plot of stream chemistry data related to four different sites in Puerto Rico (Quebrada Cuenca 1 - Bisley, Quebrada Cuenca 2 - Bisley, Quebrada Cuenca 3 - Bisley, and Rio Mameyes Puente Roto), highlighting the time when Hurricane Hugo hit in 1989.

## Load in the packages

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(janitor)
library(lubridate)
```

## Read in the four files needed to recreate the plot, cleaning the names for each one

```{r}
Q1 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca1-Bisley.csv")) %>%
  clean_names()

Q2 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca2-Bisley.csv")) %>%
  clean_names()

Q3 <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "QuebradaCuenca3-Bisley.csv")) %>%
  clean_names()

MPR <- read_csv(here("/courses", "EDS214", "equipo-nimbus", "data", "raw-data", "RioMameyesPuenteRoto.csv")) %>%
  clean_names()
```

## Combine datasets by a full join

```{r}
joined_df <- Q1 %>% # They share all of the same columns, so columns not specified
  full_join(Q2) %>%
  full_join(Q3) %>%
  full_join(MPR)
```

## Tidy the data and begin to make it ready for our analysis

```{r}
pivot_df_no3_k <- joined_df %>%
  select(sample_id, sample_date, no3_n, k) %>% # Select only necessary columns
  filter(lubridate::year(sample_date) %in% 1989:1995) %>% # Only have data from 1989-1995
  mutate(month_date = lubridate::floor_date(sample_date, "month")) %>% # Make data from the day level to the month level
  pivot_longer(cols = c(no3_n, k), names_to = "chemicals", values_to = "amounts") # Make the two chemical names in a column and values in another

k_correct_dates <- pivot_df_no3_k %>%
  filter(chemicals == "k") %>% # Select only k
  mutate(amounts = replace(amounts, lubridate::year(sample_date) >= 1993, as.numeric("NA"))) # Make anything after 1993 NAs so no values plot, to roughly match the plot we are recreating
  
pivot_df_no3_no_k <- pivot_df_no3_k %>%
  filter(chemicals == "no3_n") # Select only no3

pivot_df_no3_k_rejoined <- full_join(pivot_df_no3_no_k, k_correct_dates) # Combine refined k data with the no3 data
```

## Add a column with the level of the data as bimonthly (every two months) rather than monthly

```{r}
pivot_df_no3_k_rejoined <- pivot_df_no3_k_rejoined %>%
  mutate(odd_month_date = month_date, .after = month_date) # Create a copy of the month column which will be filled with bimonthly dates

# Converting to merge even months into odd months
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "02-01", "01-01")
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "04-01", "03-01")
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "06-01", "05-01")
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "08-01", "07-01")
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "10-01", "09-01")
pivot_df_no3_k_rejoined$odd_month_date <- str_replace_all(pivot_df_no3_k_rejoined$odd_month_date, "12-01", "11-01")

pivot_df_no3_k_rejoined$odd_month_date <- as.Date(pivot_df_no3_k_rejoined$odd_month_date) # Make it a Date variable again
```

## Create a summary table with means by month, chemical, and site

```{r}
summary_df <- pivot_df_no3_k_rejoined %>%
  group_by(sample_id, odd_month_date, chemicals) %>%
  summarize(means_by_month = mean(amounts, na.rm = TRUE))
```

## Create a plot of the two chemicals of interest (K and NO3) with colors by site

```{r}
# Define positions of vline for Hugo
date_vline <- as.Date(c("1989-09-01"))    
date_vline_position <- which(summary_df$odd_month_date %in% date_vline)

# Create the plot
recreated_plot <- ggplot(data = summary_df, aes(x = odd_month_date, 
                              y = means_by_month)) +
  geom_line(aes(color = sample_id)) +
  facet_wrap(~chemicals,
             scales = "free_y",
             ncol = 1,
             strip.position = "left", 
             labeller = as_labeller(c(k = "K mg l^-1", 
                                      no3_n = "NO3-N ug l^-1"))) +
  scale_color_manual(values = c("darkseagreen3", "bisque3", "darkgreen", "deepskyblue3")) +
  labs(title = "Bimonthly Chemical Means by Year", x = "Years", y = "Bimonthly Chemical Means", color = "") +
  theme_light() +
  geom_vline(xintercept = as.numeric(summary_df$odd_month_date[date_vline_position]), 
             color = "dark grey",
             linetype = "dashed") + # Add the line
  theme(plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank(), # Get rid of vertical grid lines 
        text=element_text(size=10,  family="Times New Roman")) # Change text size and font

recreated_plot
```

## Save the plot to the plots folder

```{r}
ggsave(plot = recreated_plot,
       here("plots", "recreated_stream_chemistry_plot.png"),
       width = 8,
       height = 8,
       dpi = 200)
```
